---
title: "PATH Forecast :: Scenarios"
author: "Planning & Regional Development"
date: "10/08/2020"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, message = FALSE, warning = FALSE, tidy.opts=list(width.cutoff=60),tidy=TRUE) # Echo is the code and include is the output  
```

#\textcolor{red}{DRAFT}

##Summary

Outputs using the PATH econmetric forecasting model and October employment inputs (moderate and protracted). Ridership and Hudson County population inputs held constant as of February 2020. 

One note of significance here is that our vendor, Oxford Economics, is projecting a decrease in *Hudson County population*, seen below. Our previous update (August) incorporated Oxford's expectation for reduced _employment_ into the PATH forecast, but those expectations remain largely unchanged between August and October; this forecast incorporates the new expectations for _population_, which puts downward pressure on the weekend projections. 

This model iteration only updates the moderate scenario. References here and in the data output files to the protracted scenario and associated data refer to the August run. 

<p>&nbsp;</p>

\pagebreak 

```{r}
setwd("~/Dropbox/Work and research/Port Authority/pathforecast")

cat("\014") # clear the console 
rm(list=ls()) 
options(scipen=999) 

library(broom) 
library(knitr) 
library(zoo) 
library(reshape2) 
library(forecast) 
library(tseries) 
library(dplyr) 
library(lubridate) 
library(doBy) 
library(mice) 
library(lmtest) 
library(tidyr) 
library(stringr) 
library(ggplot2) 
library(scales) 
library(openxlsx) 
library(writexl)
library(formatR) 
library(stats) 

start = "2004-01-01" 
```


```{r, include=TRUE, echo=TRUE}
end = "2020-02-01" 
end_and_one = "2020-03-01" 
end_cpi = "2020-10-01" 
```


```{r}
extra = as.Date(end_and_one)-as.Date(end) 
future = "2035-12-31" 

elapsed_months <- function(end_date, start_date) {
  ed <- as.POSIXlt(end_date) 
  sd <- as.POSIXlt(start_date) 
  12 * (ed$year - sd$year) + (ed$mon - sd$mon) 
}
horizon = elapsed_months(future,start)+1 
forec_horizon = elapsed_months(future,end) 
```


```{r}
# Import quarterly variables and convert quarterly to monthly. 
jobs = read.csv("./input data/econ_vars_quar 2020_10.csv") 

jobs1 = jobs#[complete.cases(jobs),] # IF QUARTERS ARE 'AVERAGE' OF RELEVANT THREE MONTHS (define 'average' later) 

jobs1$Month = as.Date(jobs1$Month, "%m/%d/%y") 

jobs1 = read.zoo(jobs1) # Converts the data frame to a time series matrix 

tt = as.yearmon(seq(start(jobs1), end(jobs1), "month")) # Makes months, different format

zm = as.data.frame(na.spline(jobs1, as.yearmon, xout = tt)) 
zm$month = seq(as.Date("2000/1/1"), as.Date("2040/10/01"), by="month") # Add date 

zm2 = subset(zm,zm$month=="2040-10-01") 
zm3 = subset(zm,zm$month=="2040-10-01") 
zm2$month="2040-11-01" 
zm3$month="2040-12-01" 
zm=rbind(zm,zm2) 
zm=rbind(zm,zm3) 

zm = zm[order(as.Date(zm$month, format="%Y-%m-%d")),] 

econ_month = zm 
names(econ_month) = tolower(names(econ_month)) 
```

```{r}
# Save economic monthly variables as: 
write.csv(econ_month,"./input data/econ_vars_months 2020_10.csv") 
```

```{r}
rm(jobs,jobs1,tt,zm,zm2,zm3) 
```


```{r}
econ_county_august = read.csv("./_misc/county_monthly_202008.csv") 
econ_county_august$month = as.Date(econ_county_august$X,format= "%m/%d/%y") 
econ_county_august$X = NULL 
econ_county_august = subset(econ_county_august,econ_county_august$month >= "2000-01-01" & econ_county_august$month <= "2035-12-31") 
econ_county_august = econ_county_august[c("month","Man.Hud.Emp.Moderate","Man.Hud.Emp.Protracted")] 
```

```{r}
econ_month = merge(econ_month,econ_county_august) 
```

```{r, echo=TRUE} 
path = read.csv("./input data/PATH_PaxCounts_2000-2009+2010-2020Apr.csv") 
path$month = as.Date(paste(path$year,str_pad(path$month, 2, pad = "0"),"01",sep="-"))
path$year = NULL 
```

```{r}
# Other data 
days = read.csv("./input data/dates_dummies.csv") 
days$month = as.Date(days$month, "%m/%d/%y") 

fare = read.csv("./input data/fare_nominal.csv") 
fare$month = as.Date(fare$month, "%m/%d/%y") 

# Merge 
path = merge(days,path, all.x = TRUE)#,by="month") 
path = merge(path,fare)#,by="month") 
path = merge(path,econ_month) 
```


```{r}
ggplot(path, aes(month)) +                    # basic graphical object
  geom_line(aes(y=Man.Hud.Emp.Moderate), colour="red") +  # first layer
  # geom_line(aes(y=Man.Hud.Emp.Protracted), colour="green") + # second layer   
  geom_line(aes(y=man_hud_2020_10), colour="blue")  # second layer 
```

```{r}
pop = data.frame(path$month,path$pop_hudson,path$pop_hudson_2020_06,path$pop_hudson_2020_10) 
names(pop) = c("month","pop_hudson_pre_covid","pop_hudson_2020_07","pop_hudson_2020_10") 
pop.melt = reshape2::melt(pop, id.var='month') 
head(pop.melt) 

ggplot(pop.melt, aes(x=month, y=value, col=variable)) + geom_line() 
#ggplot(pop.melt, aes(month)) +                    # basic graphical object
#  geom_line(aes(y=pop_hudson_pre_covid), colour="red") +  # first layer
#  geom_line(aes(y=pop_hudson_2020_07), colour="green") + # second layer   
#  geom_line(aes(y=pop_hudson_2020_10), colour="blue")  # second layer 
ggsave(
  "./plots/Hudson_Population_2020_10.png",
  plot = last_plot(),
  scale = 1,
  width = 20,
  height = 12,
  units = c("cm"),
  dpi = 300)
```


```{r, echo=TRUE}
path$cpi_base = path[path$month==end_cpi, #_cpi, 
                     "cpi_2020_10"]# "cpi_2020_10"]  
path$real_farefare = ifelse(path$month <= end_cpi, #_cpi
                            path$fare_nominal * path$cpi_base / path$cpi_2020_10, max(path$fare_nominal)) #path$cpi_2020_10, max(path$fare_nominal)) 
path$cpi_base = NULL 
```


```{r, echo=TRUE}
before = subset(path, path$month<=end & path$month>="2004-01-01") #before = head(path, 218)
after = subset(path, path$month>end) #after = tail(path, 250) 
```


```{r,  echo=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=40)}
### WEEKDAYS 
oldreg=as.matrix(data.frame(before$man_hud_2020_10, #Man.Hud.Emp.Moderate,
                            before$dummy_2,  before$dummy_3,  before$dummy_4, 
                            before$dummy_5, before$dummy_6, before$dummy_7, 
                            before$dummy_8, before$dummy_9, 
                            before$dummy_10, before$dummy_11, before$dummy_12, 
                            before$dum_911_base, before$supersandy,  before$real_farefare)) 
newreg=as.matrix(data.frame(after$man_hud_2020_10, #Man.Hud.Emp.Moderate,
                            after$dummy_2, after$dummy_3, after$dummy_4, after$dummy_5,  
                            after$dummy_6, after$dummy_7,  after$dummy_8,  after$dummy_9,  
                            after$dummy_10, after$dummy_11, after$dummy_12,
                            after$dum_911_base, after$supersandy, after$real_farefare))
```


```{r}
### SATURDAY & SUNDAY 
oldregsat=as.matrix(data.frame(before$pop_hudson,# _2020_10, #_2020_10, 
                            before$dummy_2, before$dummy_3, before$dummy_4, before$dummy_5,
                            before$dummy_6, before$dummy_7, before$dummy_8, before$dummy_9,                             before$dummy_10, before$dummy_11, before$dummy_12, 
                            before$dum_911_base, 
                            before$supersandy, before$end_close, before$real_farefare))
newregsat=as.matrix(data.frame(after$pop_hudson,# _2020_10, #_2020_10,  
                            after$dummy_2, after$dummy_3, after$dummy_4, after$dummy_5, 
                            after$dummy_6, after$dummy_7, after$dummy_8, after$dummy_9, 
                            after$dummy_10, after$dummy_11, after$dummy_12, 
                            after$dum_911_base, 
                            after$supersandy, after$end_close, after$real_farefare))
```


```{r}
#Code for scenarios
oldreg_pess = as.matrix(data.frame(before$Man.Hud.Emp.Protracted, 
                            before$dummy_2,  before$dummy_3,  before$dummy_4, 
                            before$dummy_5, before$dummy_6, before$dummy_7, 
                            before$dummy_8, before$dummy_9, 
                            before$dummy_10, before$dummy_11, before$dummy_12, 
                            before$dum_911_base, before$supersandy,  before$real_farefare)) 
newreg_pess=as.matrix(data.frame(after$Man.Hud.Emp.Protracted, 
                            after$dummy_2, after$dummy_3, after$dummy_4, after$dummy_5,  
                            after$dummy_6, after$dummy_7,  after$dummy_8,  after$dummy_9,  
                            after$dummy_10, after$dummy_11, after$dummy_12,
                            after$dum_911_base, after$supersandy, after$real_farefare))
```


##\textcolor{red}{Models are trained below}
###\textcolor{red}{Weekday:}

```{r, echo=TRUE}
fit = arima(ts(before$avg_wkdayholminor_tstile),xreg = oldreg, order=c(0,0,1), include.mean=T)
```

###\textcolor{red}{Saturday:}

```{r, echo=TRUE}
fitsat = arima(ts(before$avg_satholmajor_tstile),xreg=oldregsat,order=c(1,1,0))
```

###\textcolor{red}{Sunday:}

```{r, echo=TRUE}
fitsun = arima(ts(before$avg_sun_tstile),xreg=oldregsat,order=c(1,1,1))
```

Scenarios: 
```{r, echo=TRUE}
#Scenarios
fit_pess = arima(ts(before$avg_wkdayholminor_tstile),xreg = oldreg_pess, order=c(0,0,1), include.mean=T) 
```


```{r, echo=TRUE}
pathpredict = predict(fit, n.ahead=forec_horizon, newxreg=newreg) # level=95 #interval = "prediction", conf.level=.9) 
pathpredictsat = predict(fitsat, n.ahead=forec_horizon, newxreg=newregsat) 
pathpredictsun = predict(fitsun, n.ahead=forec_horizon, newxreg=newregsat) 
pathpredict_pess = predict(fit_pess, n.ahead=forec_horizon, newxreg=newreg_pess) 
```

```{r, echo=TRUE}
pathpredict_by_month = as.data.frame(cbind(pathpredict$pred,pathpredictsat$pred,pathpredictsun$pred,pathpredict_pess$pred)); names(pathpredict_by_month) = c("avg_wkdayholminor_tstile","avg_satholmajor_tstile","avg_sun_tstile","pess_wkdayholminor") 
pathpredict_by_month$month = seq(as.Date(end)+extra,as.Date(future),by="mon") 
```

```{r}
# Add old data back to the pile. Limit it to data from the current year. 
before_mini = subset(before,year(before$month)>=2020 & before$month<=end) 
before_mini = before_mini %>% select(avg_wkdayholminor_tstile, avg_satholmajor_tstile, avg_sun_tstile, month)
before_mini$pess_wkdayholminor = before_mini$avg_wkdayholminor_tstile
pathpredict_by_month = rbind(pathpredict_by_month, before_mini) 
```

```{r}
# Now multiply by number of days per month ... 
pathpredict_by_month = merge(pathpredict_by_month,days) 
pathpredict_by_month$sum_wkdayholminor = pathpredict_by_month$avg_wkdayholminor_tstile*pathpredict_by_month$num_wkdayholminor
pathpredict_by_month$sum_satholmajor = pathpredict_by_month$avg_satholmajor_tstile*pathpredict_by_month$num_satholmajor
pathpredict_by_month$sum_sun = pathpredict_by_month$avg_sun_tstile*pathpredict_by_month$num_sun
pathpredict_by_month$sum_wkday_pess =  pathpredict_by_month$pess_wkdayholminor*pathpredict_by_month$num_wkdayholminor
```


```{r, echo=TRUE}
# Annual 
pathpredict_by_month$year = year(pathpredict_by_month$month) 
pathpredict_year = summaryBy(sum_wkdayholminor + sum_satholmajor + sum_sun + sum_wkday_pess ~ year, data = pathpredict_by_month, FUN = sum); names(pathpredict_year) = c("year","base_wkday", "saturday", "sunday","pess_wkday") 
pathpredict_year$base_total = pathpredict_year$base_wkday + pathpredict_year$saturday + pathpredict_year$sunday 
pathpredict_year$pess_total = pathpredict_year$pess_wkday + pathpredict_year$saturday + pathpredict_year$sunday 
pathpredict_by_month$year = NULL 
```

```{r}
pathpredict_year = pathpredict_year[c(1,2,3,4,6,5,7)] 
#pathpredict_year = pathpredict_year[c(1,2,3,4,7,5,8,6,9)]
```


```{r}
out1 = tidy(fit) 
out2 = glance(fit) 
out2.5 = accuracy(fit) 
out3 = tidy(fitsat) 
out4 = glance(fitsat)
out4.5 = accuracy(fitsat) 
out5 = tidy(fitsun) 
out6 = glance(fitsun) 
out6.5 = accuracy(fitsun) 
```


```{r, echo=TRUE}
resids = as.data.frame(cbind(as.vector(resid(fit)),as.vector(resid(fitsat)),as.vector(resid(fitsun)))); names(resids) = c("Weekday_residuals", "Saturday_residuals", "Sunday_residuals")
```


\pagebreak 

##Output

```{r}
pathpredict_by_month$weekdays = NULL 
pathpredict_by_month$saturdays = NULL 
pathpredict_by_month$sundays = NULL 
pathpredict_by_month$year = NULL 
```

```{r}
tail(pathpredict_year) 
```


Save everything as: 
```{r, echo=TRUE}
sheets = list(Data = path, Monthly_Output = pathpredict_by_month, Annual_Output = pathpredict_year, Residuals = resids) 
write_xlsx(sheets, "./output data/Output Scenarios 2020-10 moderate.xlsx") # This exports and names the file. 
```

###Monthly output included in Excel file within output folder. 

\pagebreak 

##Modeling statistics and diagnostics

###Weekday

```{r}
kable(out1,caption="Weekday Coefficients") 
kable(out2,caption="Weekday Diagnostics")
kable(out2.5,caption="Weekday Additional Diagnostics")
```

\pagebreak 

###Saturday

```{r}
kable(out3,caption="Saturday Coefficients") 
kable(out4,caption="Saturday Diagnostics")
kable(out4.5,caption="Saturday Additional Diagnostics")
```

\pagebreak

###Sunday

```{r}
kable(out5,caption="Sunday Coefficients") 
kable(out6,caption="Sunday Diagnostics")
kable(out6.5,caption="Sunday Additional Diagnostics")
```

```{r}
# plot the first curve by calling plot() function
# First curve is plotted
{plot(econ_month$month,econ_month$man_hud, type="o", col="blue", pch="o", lty=1, ylim=c(1500,4500), ylab="Man+Hud Employment")
points(econ_month$month,econ_month$man_hud_2020_10, col="red", pch="*")
points(econ_month$month,econ_month$Man.Hud.Emp.Protracted, col="green", pch="o")}
#lines(econ_month$month,econ_month$Man.Hud.Emp.Moderate, col="red",lty=2)
```

