---
title: "PATH 2020-06"
author: "Planning & Regional Development"
date: "5/29/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Copy and move input files - input variables and dates. DONE. 
Incorporate quarter-to-month interpolation. DONE. 
Bind dates, ridership, monthly economics, other. DONE. 
Add and update diagnostics. 
Automate real fare calculation. 
Confirm represetns 2020 Q1 (?) forecast outputs. 
Package outputs (inputs, forecast and diagnostics). 
Clean code. 
Add narrative. 
Package (zip?) and share. 

##1. Setup setup setup.

```{r}
setwd("~/Dropbox/Work and research/Port Authority/pathforecast")

cat("\014") # clear the console 
rm(list=ls()) 
options(scipen=999) 

library(broom) 
library(knitr) 
library(zoo) 
library(reshape2) 
library(forecast) 
library(tseries) 
library(dplyr) 
library(lubridate) 
library(doBy) 
library(mice) 
library(lmtest) 
library(tidyr) 
library(zoo) 
#library(yardstick) 

set.seed(101) 

start = "2004-01-01" 
end = "2020-02-01" #"2019-12-01" 
end_and_one = "2020-03-01" #"2020-01-01" 
extra = as.Date(end_and_one)-as.Date(end) 
future = "2040-12-31" 

elapsed_months <- function(end_date, start_date) {
  ed <- as.POSIXlt(end_date) 
  sd <- as.POSIXlt(start_date) 
  12 * (ed$year - sd$year) + (ed$mon - sd$mon) 
}
horizon = elapsed_months(future,start)+1 
forec_horizon = elapsed_months(future,end) 
forec_horizon
```


##2. Interpolation: convert quarterly data to monthly. 

```{r}
jobs = read.csv("./econ_vars_quar 2020_06.csv") 
jobs1 = jobs#[complete.cases(jobs),] 

# IF QUARTERS ARE 'AVERAGE' OF RELEVANT THREE MONTHS (define 'average' later) 
jobs1$year=NULL 
jobs1$quarter=NULL 
jobs1$Indicator=NULL 

jobs1$Month = as.Date(jobs1$Month, "%m/%d/%y") 

jobs2 = jobs1 
jobs2$quarter2 = NULL 
jobs2 = read.zoo(jobs2) # Converts the data frame to a time series matrix 

tt = as.yearmon(seq(start(jobs2), end(jobs2), "month")) # Makes months, different format (unsure why needed) 
jobs2$Indicator = NULL 

zm = as.data.frame(na.spline(jobs2, as.yearmon, xout = tt)) 
zm$month_ = seq(as.Date("1996/1/1"), as.Date("2035/10/01"), by="month") # Add date 

zm2 = subset(zm,zm$month_=="2035-10-01") 
zm3 = subset(zm,zm$month_=="2035-10-01") 
zm2$month_="2035-11-01" 
zm3$month_="2035-12-01" 
zm=rbind(zm,zm2) 
zm=rbind(zm,zm3) 
zm = zm[order(as.Date(zm$month_, format="%Y-%m-%d")),] 

write.csv(zm,"./econ_vars_months 2020_06.csv") 
jobs_month = zm 
names(jobs_month) = tolower(names(jobs_month)) 

rm(jobs,jobs1,jobs2,tt,zm,zm2,zm3) 
```

##3. Load data load data.

```{r}
days = read.csv("./Dates_dummies.csv") 
#; names(days) = c("month","weekdays","saturdays","sundays") 
days$month = as.Date(days$month, "%m/%d/%y") 

path = read.csv("./PATH input 2020_06.csv") 
path$month = as.Date(days$month, "%m/%d/%y") 

other = read.csv("./other.csv") 
other$month = as.Date(other$month, "%m/%d/%y") 

path = merge(days,path)#,by="month") 
path = merge(path,other)#,by="month") 
#path$month = as.Date(days$month,format="%m/%d/%y") 
```


##4.Prepare for model.

```{r}
#before = subset(path,path$month<=end & path$month>="2002-01-01") 
before = head(path,218) 
#after = subset(path,path$month>end) 
after = tail(path,250) 
summary(after) 

### WEEKDAYS 
oldreg=as.matrix(data.frame(before$man_hud_KEEP,                             before$dummy_2,before$dummy_3,before$dummy_4,before$dummy_5,before$dum_911_base,
                      before$dummy_6,before$dummy_7,before$dummy_8,before$dummy_9,before$dummy_10,before$dummy_11,before$dummy_12,
                            before$supersandy, before$real_fare_q1_KEEP)) #real_fare_q4
newreg=as.matrix(data.frame(after$man_hud_KEEP, 
                            after$dummy_2,after$dummy_3,after$dummy_4,after$dummy_5, after$dum_911_base,
                            after$dummy_6, after$dummy_7, after$dummy_8, after$dummy_9, after$dummy_10, after$dummy_11, after$dummy_12,
                            after$supersandy, after$real_fare_q1_KEEP))

### SATURDAY & SUNDAY 
oldregsat=as.matrix(data.frame(before$pop_hudson_KEEP,before$dummy_2, before$dummy_3, before$dummy_4,before$dum_911_base,
                            before$dummy_5, before$dummy_6, before$dummy_7, before$dummy_8, before$dummy_9, before$dummy_10, before$dummy_11, 
                            before$dummy_12,before$supersandy, before$end_close,
                            before$real_fare_q1_KEEP))
newregsat=as.matrix(data.frame(after$pop_hudson_KEEP,after$dummy_2, after$dummy_3, after$dummy_4,after$dum_911_base,
                            after$dummy_5, after$dummy_6, after$dummy_7, after$dummy_8, after$dummy_9, after$dummy_10, after$dummy_11, 
                            after$dummy_12,after$supersandy, after$end_close,
                            after$real_fare_q1_KEEP)) 

#t.test(before$sun,before$real_fare) 
```

##5. Models and forecasts. 
###5a. Fit models (estimate equations). 
###Weekday
```{r}
summary(before) 
fit = arima(ts(before$avg_wkdayholminor_tstile),xreg = oldreg, order=c(0,0,1), include.mean=T)#method="CSS")# as of 2018-09 (1,0,1) before that (1,1,2)) 
#fit = arima(ts(before$avg_wkdayholminor_tstile),xreg = oldreg, order=c(0,0,1), include.mean=T)# as of 2018-09 (1,0,1) before that (1,1,2)) 
  #fit = auto.arima(ts(before$week),xreg=oldreg, ic="aic", trace=TRUE, allowdrift=FALSE)#,lambda=0, seasonal=TRUE)# 
```

###Saturday
```{r}
fitsat = arima(ts(before$avg_satholmajor_tstile),xreg=oldregsat,order=c(1,1,0))# as of 2018-09 (1,1,0) before that (1,0,2), method="CSS")# 
#fitsat = arima(ts(before$avg_satholmajor_tstile),xreg=oldregsat,order=c(1,1,0))# as of 2018-09 (1,1,0) before that (1,0,2), method="CSS")# 
  #fitsat = auto.arima(ts(before$sat_mice),xreg=oldregsat, ic="aic", trace=TRUE, allowdrift=FALSE ,lambda=0, seasonal=TRUE)# 
```

###Sunday
```{r}
fitsun = arima(ts(before$avg_sun_tstile),xreg=oldregsat,order=c(1,1,1))# as of 2018-09 (1,1,1) before that (1,0,2)) 
#fitsun = arima(ts(before$avg_sun_tstile),xreg=oldregsat,order=c(1,1,1))# as of 2018-09 (1,1,1) before that (1,0,2)) 
  #fitsun = auto.arima(ts(before$sun_mice),xreg=oldregsat, ic="aic", trace=TRUE, allowdrift=FALSE,lambda=0, seasonal=TRUE) 
```

###5b. Predict (forecast).
```{r}
pathpredict = predict(fit, n.ahead=forec_horizon, newxreg=newreg, level=95)#interval = "prediction", conf.level=.9) # predict 
pathpredictsat = predict(fitsat, n.ahead=forec_horizon, newxreg=newregsat) # predict 
pathpredictsun = predict(fitsun, n.ahead=forec_horizon, newxreg=newregsat) # predict 

```

##6. Clean and consolidate results for export. 

```{r}
pathpredict_by_month = as.data.frame(cbind(pathpredict$pred,pathpredictsat$pred,pathpredictsun$pred)); names(pathpredict_by_month) = c("week_avg","sat_avg","sun_avg") 
head(pathpredict_by_month,3) 
end 
future 

pathpredict_by_month$month = seq(as.Date(end)+extra,as.Date(future),by="mon") 

## Add old stuff (January 2017, for example) back to the pile. 
before_mini = data.frame((before$week*before$weekdays),(before$sat*before$saturdays),(before$sun*before$sundays),before$month) 
  names(before_mini) = c("week","sat","sun","month") 

## Now multiply by number of days per month ... 
pathpredict_by_month = merge(pathpredict_by_month,days) 
pathpredict_by_month$week = pathpredict_by_month$week_avg*pathpredict_by_month$weekdays 
pathpredict_by_month$sat = pathpredict_by_month$sat_avg*pathpredict_by_month$saturdays 
pathpredict_by_month$sun = pathpredict_by_month$sun_avg*pathpredict_by_month$sundays 


pathpredict_mini = data.frame(pathpredict_by_month$month,pathpredict_by_month$week,pathpredict_by_month$sat,pathpredict_by_month$sun) 
  names(pathpredict_mini) = c("month","week","sat","sun") 

```

##7. Model diagnostics. 

```{r}
  
out1 = tidy(fit) 
  #out2 = tidy(glance(fit)) ## why is this crashing my program? 
out2 = glance(fit) 
out1  
out2 
out3 = tidy(fitsat) 
out4 = glance(fitsat)
out3
out4 
out5 = tidy(fitsun) 
out6 = glance(fitsun)
out5 
out6 

accuracy(fit) 
accuracy(fit)[,'MAPE'] 

pathpredict_month_backup = pathpredict_by_month 
pathpredict_month = rbind(before_mini,pathpredict_mini) #meh. figure this out later


pathpredict_month$year = year(pathpredict_month$month) 

pathpredict_year = summaryBy(week + sat + sun ~ year, data = pathpredict_month, FUN = sum); names(pathpredict_year) = c("year","week","sat","sun")
pathpredict_year$total = pathpredict_year$week + pathpredict_year$sat + pathpredict_year$sun 
pathpredict_month$year = NULL 
years = pathpredict_year 
pathpredict_year[14,5] = 76565451 
pathpredict_year[15,5] = 78517120 


resids = as.data.frame(cbind(as.vector(resid(fit)),as.vector(resid(fitsat)),as.vector(resid(fitsun)))) 
```

##8. Save (export).

```{r}
#write.csv(pathpredict_by_month,"./PATH forecast products/PATH forecast output/PATH q2/PATH month_ 20190506.csv")
 write.csv(pathpredict_month_backup, "./PATH output test 20200529.csv") 
#  write.csv(fitted(fit), "./PA PATH output & viz/PATH fitted _week 2020q1.csv") 
#  write.csv(fitted(fitsat), "./PA PATH output & viz/PATH fitted _sat 2020q1.csv") 
#  write.csv(fitted(fitsun), "./PA PATH output & viz/PATH fitted _sun 2020q1.csv") 
#  write.csv(resids, "./PATH forecast products/PATH forecast output/PATH q2/PATH residuals _nodummy 2019q2.csv")

#write.csv(pathpredict_year,"./PA PATH output & viz/PATH ANNUAL_Q4 FARE TRNSTL RIDERS PESS.csv") 
#  write.csv(pathpredict_by_month, "./PA PATH output & viz/PATH MONTH_2020Q1.csv") 

```

Did the final year of output total 100,347,550? 
